Ubiquitous Digital Technologies and Spatial Structure; an update

### 2020-12-02 paper/2_resubmission/tranos_ioannides.tex
minor changes

Kolko {[}22{]} uses internet diffusion data and identifies a clear
complementary link between internet usage and city size. Interestingly,
he identifies higher internet domain densities in remote cities which
indicated a substitution effect of the internet for longer-distance
non-electronic communications. His results are consistent for different
measures of internet diffusion (internet domain density and internet
take-up). Sinai and Waldfogel {[}23{]} approach the same question from
the consumers' point of view and find support for complementarity
between internet and cities. They study the link between market size and
locally targeted online content and find that more online local content
in larger markets. However, they also obtain evidence for a substitution
effect: they find that holding constant local online content constant,
market size has a negative effect on individual connectivity. Forman et
al.~{[}24{]} examined whether commercial internet adoption is higher in
cities than in rural areas. While the former would indicate a
complementarity between internet adoption and cities, the latter would
reflect a complementary relation, according to which the internet is
used as a means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this was not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

@@

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the Zipf
coefficient for a broad sample of countries over time. Zipf coefficient
is one of the most widely used measures of spatial dispersion with
numerous applications in urban economics and economic geography ({[}see
for example {[}36{]}; {[}37{]}; {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]};
{[}42{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations. Skipping details
pertaining of the suitability of the approach, which may be found in
Ioannides et al.~{[}18{]} and Tranos and Ioannides {[}19{]}, we proceed
with the estimation of the logarithmic form of:

@@

This leads to an estimate of \(\zeta\), known as the Zipf coefficient,
but more appropriately the Pareto exponent, when \(\zeta\) is not
estimated to be near \(1\). While much of the literature focuses on the
Zipf coefficient being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\):

@@

The estimation of equation \ref{eq2} has typically been performed by
Ordinary Least Squares (OLS). Gabaix and Ioannides {[}43{]} discuss the
downward bias of estimates of equation \ref{eq2} using OLS on small
samples. Gabaix and Ibragimov {[}44{]} propose a practical remedy to
correct this bias, which we do adopt in this paper: instead of using the
\(log\) of \(rank\) of a city \(i\) in a country \(c\) in year \(t\),
they propose to use the \(log(rank-0.5)\), which has indeed been widely
adopted. Researchers working in this area must contend with definitional
differences as well as differences in availability of different kinds of
data sources. Definitions of cities differ across countries, for
political, administrative and legal reasons\footnote{See {[}45{]}, Ch.
  8, for issues and pitfalls associated with different definitions.}. In
contrast to the data drawn from Thomas Brinkhoff's City Population
project {[}46{]}\footnote{\url{http://www.citypopulation.de}}, which
were used by Soo {[}47{]}, Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, the multi-country analysis in the present paper is
based on the annual population data for urban agglomerations with
\(300,000\) inhabitants or more from the Department of Economic and
Social Affairs of the United Nations {[}9{]}. Despite some criticism
about the consistency of the urban agglomeration definitions across
different countries {[}48{]}, this is the only available source for
yearly, multi-country population data for urban agglomerations
{[}49--51{]}. The results based on these data are reported and discussed
in \protect\hyperlink{sec4}{Section 4}.

@@

\caption{\label{tab:unnamed-chunk-2}Zipf coefficients\label{zipf}}

@@

Zipf coef. & 2000 & 2001 & 2002 & 2003 & 2004 & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018\\

@@

As Table \ref{cor} indicates, although there some rather strong
correlations among these variables, mobile phone penetration appears to
have a distinct character from its fixed phone counterpart: their
correlation coefficient is only \(0.445\). This probably highlights the
different composition of the population or infrastructure development
patterns in the developing world, where mobile telephony helped overcome
the lack of fixed line infrastructure and mobile phone networks are also
used as the main way to use the internet {[}52,53{]}.

\begin{table}[!htbp] \centering 
  \caption{Descriptive statistics for the multi-country model\label{desc.global}} 
  \label{} 
\footnotesize 
\begin{tabular}{@{\extracolsep{1pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Zipf coefficient & 950 & $-$1.9 & $-$1.1 & 0.2 & $-$0.7 \\ 
Zipf inv. sq. error\textsuperscript{a} & 950 & 0.01 & 0.1 & 0.05 & 0.2 \\ 
Population density & 931 & 2.5 & 126.2 & 182.1 & 1,239.6 \\ 
Government expenditure (\% GDP) & 913 & 1.0 & 14.7 & 4.9 & 30.0 \\ 
Trade (\% GDP) & 914 & 19.1 & 65.3 & 33.9 & 220.4 \\ 
GDP growth & 934 & $-$33.1 & 4.0 & 4.2 & 54.2 \\ 
GDP per capita (log) & 892 & 630.7 & 18,049.2 & 15,432.9 & 61,391.4 \\ 
Female labour force (\%) & 950 & 7.9 & 37.8 & 11.9 & 55.2 \\ 
Population & 950 & 5,122,493 & 116,073,037.0 & 244,323,372.0 & 1,392,730,000 \\ 
Non agriculture value added (\% GDP) & 930 & 58.8 & 90.3 & 8.2 & 99.4 \\ 
Internet users per 100 hab. & 916 & 0.01 & 32.5 & 28.2 & 96.0 \\ 
Broadband users per 100 hab. & 824 & 0.0 & 8.8 & 11.2 & 44.8 \\ 
Mobile phone users per 100 hab. & 949 & 0.0 & 72.9 & 46.4 & 191.0 \\ 
Fixed phone users per 100 hab. & 946 & 0.0 & 19.8 & 18.6 & 68.4 \\ 
\hline \\[-1.8ex] 
\multicolumn{6}{l}{\textsuperscript{a}This is the inverse squared standard error of the estimated Zipf, coefficient, which is used for weighting} \\ 
\multicolumn{6}{l}{the observations for the estimation, of equation 3. See Section 4} \\ 
\end{tabular} 
\end{table}

@@

The availability of a panel dataset for city sizes across countries
enables us to use country fixed effects, which can address potential
endogeneity issues related to unobserved country specific
characteristics of city size distributions. However, such a strategy
does not address potential simultaneity issues. Simply put, internet
penetration might be affected by spatial structure, as reflected in Zipf
coefficients, or both internet penetration and spatial structure might
be jointly determined by a third variable. E.g., if a country already
has a dispersed spatial structure, internet is particularly suitable in
facilitating communication. Potential endogeneity in our specification
will prevent us from being able to determine the causal impact of
internet and digital communication technologies usage on spatial
structure, which is the main aim of this paper. In order to address this
problem, we will adopt an instrumented variable strategy. Table
\ref{desc.global} also includes the descriptive statistics for the
instrumental variable we are using and (\(female\:labour\:force\)), will
be discussed in \protect\hyperlink{sec4}{Section 4}.

@@

In order to capture the micro-dynamics of the urban systems in the two
case studies, we follow Batty {[}54{]} and Havlin {[}55{]} and focus on
the difference in ranks for individual cities during the study
period.\footnote{Changes in ranking of individuals has also been used
  elsewhere in economics because of its robustness properties. C.f.
  Chetty and Hendren {[}56{]}.} Contrary to their approach, we are
interested in the real difference in rankings rather than in absolute
differences, in order to capture whether or a city improves its ranking
within the urban system during the study period. We then test whether
our internet-related variable has an effect on the change in ranking.
Hence, we define the left-hand side (LHS) variable of equation \ref{eq4}
as follows:

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the Zipf coefficient, as estimated according to
the Gabaix and Ibragimov {[}44{]} correction. The main explanatory
variables of interest are provided at the top of Table \ref{ols.global};
namely internet, broadband, mobile and fixed telephony per \(100\)
habitants, expressed in natural logarithms. The main variables of
interest are introduced successively on their own in the regressions
reported in Table \ref{ols.global}. All regressions include country
fixed effects to control for unobserved heterogeneity and a time trend.
In addition, the observations are weighted with the inverse squared
standard error of the estimated Zipf coefficient (see Table
\ref{desc.global}) to address potential noise that is carried over from
the first part of our identification strategy. In regards to the
interpretation of the estimated coefficients, given that the Zipf
coefficient has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

\begin{table}[!htbp] \centering 
  \caption{OLS estimation of equation (3)\label{ols.global}} 
  \label{} 
\small 
\begin{tabular}{@{\extracolsep{1pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Zipf Coefficient 2000-18} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\ 
  & (0.0001) &  &  &  \\ 
  Broadband users per 100 hab. (log) &  & $-$0.001$^{**}$ &  &  \\ 
  &  & (0.0002) &  &  \\ 
  Mobile phone users per 100 hab. (log) &  &  & 0.0003 &  \\ 
  &  &  & (0.0003) &  \\ 
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.00003 \\ 
  &  &  &  & (0.0001) \\ 
  Population density (log) & $-$1.378 & $-$1.060 & $-$1.505 & $-$1.892 \\ 
  & (1.870) & (1.985) & (1.896) & (2.007) \\ 
  Government expenditure (% GDP) & 0.0002 & 0.003$^{***}$ & 0.001 & 0.001 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  Trade (\% of GDP) & 0.0002$^{***}$ & $-$0.0001 & 0.0001 & 0.0001 \\ 
  & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\ 
  Non agriculture value added (\% GDP) & 0.0001 & 0.001 & $-$0.0001 & $-$0.0002 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  GDP growth & 0.001$^{***}$ & 0.001$^{***}$ & 0.001$^{**}$ & 0.001$^{**}$ \\ 
  & (0.0003) & (0.0004) & (0.0004) & (0.0004) \\ 
  GDP per capita (log) & $-$0.020$^{***}$ & $-$0.023$^{***}$ & $-$0.019$^{***}$ & $-$0.016$^{**}$ \\ 
  & (0.007) & (0.006) & (0.007) & (0.008) \\ 
  Population (log) & 1.300 & 0.955 & 1.390 & 1.785 \\ 
  & (1.873) & (1.985) & (1.899) & (2.011) \\ 
  Constant & $-$18.795 & $-$13.969 & $-$19.948 & $-$25.516 \\ 
  & (26.314) & (27.886) & (26.670) & (28.261) \\ 
 \hline \\[-1.8ex] 
Country fixed effects & Yes & Yes & Yes & Yes \\ 
Yearly fixed effects & Yes & Yes & Yes & Yes \\ 
Observations & 844 & 757 & 865 & 867 \\ 
Adjusted R$^{2}$ & 0.987 & 0.988 & 0.986 & 0.986 \\ 
Residual Std. Error & 0.516 & 0.497 & 0.540 & 0.540 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{4}{r}{Robust Std. Errors in parenthesis} \\ 
\end{tabular} 
\end{table}

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the Zipf coefficient, might be affected by another source,
which also affects internet penetration. For instance, economic
development might affect the concentration of population in large cities
and at the same time enable more people to go online. If we do not
address this issue, the coefficient for the main variable of interest
will capture potential effects that internet penetration has on spatial
structure, but also potential reverse causality effects that spatial
structure might generate on internet penetration. To overcome this
potential problem, Table \ref{2sls.global} reports estimates of equation
\ref{eq3} using two-stage least squares (2SLS) with instrumental
variables (IVs). The latter are variables which are correlated with our
endogenous variables, but do not influence current spatial structure.
Such an approach will enable us to estimate the causal effect -- if any
-- of the internet and digital communications on spatial structure. At a
first stage, our endogenous variables are regressed against the IVs.
Then, the predicted values of the endogenous variables based on the IVs
and the other control variables are used instead of the endogenous
variable to estimate eq. \ref{eq3}. A significant effect will verify the
causal impact of the internet and digital communication usage on spatial
structure.

>>>>

Kolko {[}22{]} uses internet diffusion data and identifies a clear
complementary link between internet usage and city size. Interestingly,
he identifies higher internet domain densities in remote cities which
indicates a substitution effect of the internet for longer-distance
non-electronic communications. His results are consistent for different
measures of internet diffusion (internet domain density and internet
take-up). Sinai and Waldfogel {[}23{]} approach the same question from
the consumers' point of view and find support for complementarity
between internet and cities. They study the link between market size and
locally targeted online content and find more online local content in
larger markets. However, they also obtain evidence for a substitution
effect: holding local online content constant, market size has a
negative effect on individual connectivity. Forman et al.~{[}24{]}
examines whether commercial internet adoption is higher in cities than
in rural areas. While the former would indicate a complementarity
between internet adoption and cities, the latter would reflect a
substitutability relation, according to which the internet is used as a
means to offset costs and lack of opportunities related to
peripheriality. Their results indicate that despite internet adoption by
firms with more than 100 employees being greater in smaller urban
agglomerations, the adoption of more sophisticated internet-based
applications is positively related with city size in \(2000\). Sohn,
Kim, and Hewings {[}25{]} compare how information technologies are
related to urban spatial structure for Chicago and Seoul. Although they
find a clear complementary link for Chicago, this is not the case for
Seoul, where information technologies contribute to a more dispersed
spatial pattern. Focusing on the municipalities in the province of
Barcelona, Pons-Novell and Viladecans-Marsal {[}26{]} find a
complementary link between individual internet take-up and off-line
commercial offerings, but their results cannot safely reject a
substitution effect. Craig, Hoang, and Kohlhase {[}27{]} focus on
internet take-up rates across US states during the period \(2000-2011\).
Their analysis provides suggestive evidence of a complementary role that
internet connectivity performs on urban living. Ioannides et
al.~{[}18{]} use country-level data on city size distributions to
examine the impact of fixed line telephony on urban structure
\(1980-2000\). Using a panel dataset of spatial dispersion measures,
they find robust evidence that an increase in the number of telephone
lines per capita encourages the spatial dispersion of population in that
they lead to a more concentrated distribution of city sizes. In
addition, by the end of their study period the internet has come into
use, but their results with internet usage is more speculative but do
show that it goes in the same direction.

@@

The present paper does not address the issue that adoption of ICT may
have different effects on urban dispersion on a national scale, that is
how far are major urban areas, that is, cities, from one another, versus
dispersion within major urban areas, that is urban sprawl. Such an
inquiry would require detailed data about patterns of how urban sprawl
may hamper or promote economic interaction, given advancements in
information technology. We think that data from the Covid-19 pandemic
would likely throw light at this issue in the future, as it is clear
that not all economic activities might continue in an unencumbered
manner; see Dingel and Neiman {[}36{]}. \color{black}

@@

The multi-country identification strategy is a two-step approach
{[}18{]}. The first step of our methodology is to estimate the
\color{blue} Pareto exponent \color{black} for a broad sample of
countries over time. \color{blue} The Pareto exponent \color{black} is
one of the most widely used measures of spatial dispersion with numerous
applications in urban economics and economic geography ({[}see for
example {[}37{]}; {[}38{]}; {[}39{]}; {[}40{]}; {[}41{]}; {[}42{]};
{[}43{]}). This an appropriate measure of dispersion because of the
extreme heterogeneity of the city size distribution and the very good
fits normally obtained with such estimations\footnote{\color{blue} We
  also replicate our analysis using other measures of dispersion in
  Appendix S1. Our results remain consistent.}. \color{black} Skipping
details pertaining of the suitability of the approach, which may be
found in Ioannides et al.~{[}18{]} and Tranos and Ioannides {[}19{]}, we
proceed with the estimation of the logarithmic form of:

@@

This leads to an estimate of \(\zeta\), known as the Zipf coefficient,
but more appropriately the Pareto exponent, a terminology we will adhere
to for the remained of the text, even when \(\zeta\) is estimated to be
near \(1\). The term Zipf coefficient should be reserved for the special
case of \(\zeta=1\). While much of the literature focuses on the Pareto
exponent being close to \(1\), several estimations of city size
distributions lead to estimates of \(\zeta\) that are not necessarily
equal to \(1\), or even near it, in which case we refer to \(\zeta\) as
the Pareto, or power law, exponent. Given that our aim here is to
estimate \(\zeta\) for a number of countries over time as a measure of
dispersion, equation \ref{eq1} describes the rank of city \(i\) in
country \(c\) in year \(t\): \color{black}

@@

The estimation of equation \ref{eq2} has typically been performed by
Ordinary Least Squares (OLS). Gabaix and Ioannides {[}44{]} discuss the
downward bias of estimates of equation \ref{eq2} using OLS on small
samples. Gabaix and Ibragimov {[}45{]} propose a practical remedy to
correct this bias, which we do adopt in this paper: instead of using the
\(log\) of \(rank\) of a city \(i\) in a country \(c\) in year \(t\),
they propose to use the \(log(rank-0.5)\), which has indeed been widely
adopted. Researchers working in this area must contend with definitional
differences as well as differences in availability of different kinds of
data sources. Definitions of cities differ across countries, for
political, administrative and legal reasons\footnote{See {[}46{]}, Ch.
  8, for issues and pitfalls associated with different definitions.}. In
contrast to the data drawn from Thomas Brinkhoff's City Population
project {[}47{]}\footnote{\url{http://www.citypopulation.de}}, which
were used by Soo {[}48{]}, Ioannides et al.~{[}18{]} and Tranos and
Ioannides {[}19{]}, the multi-country analysis in the present paper is
based on the annual population data for urban agglomerations with
\(300,000\) inhabitants or more from the Department of Economic and
Social Affairs of the United Nations {[}9{]}. Despite some criticism
about the consistency of the urban agglomeration definitions across
different countries {[}49{]}, this is the only available source for
yearly, multi-country population data for urban agglomerations
{[}50--52{]}. The results based on these data are reported and discussed
in \protect\hyperlink{sec4}{Section 4}.

@@

\caption{\label{tab:unnamed-chunk-2}Pareto exponent\label{zipf}}

@@

Countries & 2000 & 2001 & 2002 & 2003 & 2004 & 2005 & 2006 & 2007 & 2008 & 2009 & 2010 & 2011 & 2012 & 2013 & 2014 & 2015 & 2016 & 2017 & 2018\\

@@

As Table \ref{cor} indicates, although there some rather strong
correlations among these variables, mobile phone penetration appears to
have a distinct character from its fixed phone counterpart: their
correlation coefficient is only \(0.445\). This probably highlights the
different composition of the population or infrastructure development
patterns in the developing world, where mobile telephony helped overcome
the lack of fixed line infrastructure and mobile phone networks are also
used as the main way to use the internet {[}53,54{]}.

\begin{table}[!htbp] \centering 
  \caption{Descriptive statistics for the multi-country model\label{desc.global}} 
  \label{} 
\footnotesize 
\begin{tabular}{@{\extracolsep{1pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St. Dev.} & \multicolumn{1}{c}{Max} \\ 
\hline \\[-1.8ex] 
Pareto exponent & 950 & $-$1.9 & $-$1.1 & 0.2 & $-$0.7 \\ 
Pareto exp. inv. sq. error\textsuperscript{a} & 950 & 0.01 & 0.1 & 0.05 & 0.2 \\ 
Population density & 931 & 2.5 & 126.2 & 182.1 & 1,239.6 \\ 
Government expenditure (\% GDP) & 913 & 1.0 & 14.7 & 4.9 & 30.0 \\ 
Trade (\% GDP) & 914 & 19.1 & 65.3 & 33.9 & 220.4 \\ 
GDP growth & 934 & $-$33.1 & 4.0 & 4.2 & 54.2 \\ 
GDP per capita (log) & 892 & 630.7 & 18,049.2 & 15,432.9 & 61,391.4 \\ 
Female labour force (\%) & 950 & 7.9 & 37.8 & 11.9 & 55.2 \\ 
Population & 950 & 5,122,493 & 116,073,037.0 & 244,323,372.0 & 1,392,730,000 \\ 
Non agriculture value added (\% GDP) & 930 & 58.8 & 90.3 & 8.2 & 99.4 \\ 
Internet users per 100 hab. & 916 & 0.01 & 32.5 & 28.2 & 96.0 \\ 
Broadband users per 100 hab. & 824 & 0.0 & 8.8 & 11.2 & 44.8 \\ 
Mobile phone users per 100 hab. & 949 & 0.0 & 72.9 & 46.4 & 191.0 \\ 
Fixed phone users per 100 hab. & 946 & 0.0 & 19.8 & 18.6 & 68.4 \\ 
\hline \\[-1.8ex] 
\multicolumn{6}{l}{\textsuperscript{a}This is the inverse squared standard error of the estimated Pareto exponent, which is used for weighting} \\ 
\multicolumn{6}{l}{the observations for the estimation of equation 3. See Section 4} \\ 
\end{tabular} 
\end{table}

@@

The availability of a panel dataset for city sizes across countries
enables us to use country fixed effects, which can address potential
endogeneity issues related to unobserved country specific
characteristics of city size distributions. However, such a strategy
does not address potential simultaneity issues. Simply put, internet
penetration might be affected by spatial structure, as reflected in
Pareto exponents, or both internet penetration and spatial structure
might be jointly determined by a third variable. E.g., if a country
already has a dispersed spatial structure, internet is particularly
suitable in facilitating communication. Potential endogeneity in our
specification will prevent us from being able to determine the causal
impact of internet and digital communication technologies usage on
spatial structure, which is the main aim of this paper. In order to
address this problem, we will adopt an instrumented variable strategy.
Table \ref{desc.global} also includes the descriptive statistics for the
instrumental variable we are using and (\(female\:labour\:force\)), will
be discussed in \protect\hyperlink{sec4}{Section 4}.

@@

In order to capture the micro-dynamics of the urban systems in the two
case studies, we follow Batty {[}55{]} and Havlin {[}56{]} and focus on
the difference in ranks for individual cities during the study
period.\footnote{Changes in ranking of individuals has also been used
  elsewhere in economics because of its robustness properties. C.f.
  Chetty and Hendren {[}57{]}.} Contrary to their approach, we are
interested in the real difference in rankings rather than in absolute
differences, in order to capture whether or a city improves its ranking
within the urban system during the study period. We then test whether
our internet-related variable has an effect on the change in ranking.
Hence, we define the left-hand side (LHS) variable of equation \ref{eq4}
as follows:

@@

This section presents the estimation results of \ref{eq3}. The LHS
variable is \(\zeta\), the \color{blue} Pareto exponent, \color{black}
as estimated according to the Gabaix and Ibragimov {[}45{]} correction.
The main explanatory variables of interest are provided at the top of
Table \ref{ols.global}; namely internet, broadband, mobile and fixed
telephony per \(100\) habitants, expressed in natural logarithms. The
main variables of interest are introduced successively on their own in
the regressions reported in Table \ref{ols.global}. All regressions
include country fixed effects to control for unobserved heterogeneity
and a time trend. In addition, the observations are weighted with the
inverse squared standard error of the estimated \color{blue} Pareto
exponent \color{black} (see Table \ref{desc.global}) to address
potential noise that is carried over from the first part of our
identification strategy. In regards to the interpretation of the
estimated coefficients, given that the \color{blue} Pareto exponent
\color{black} has entered the regression not as an absolute value, but
instead as a real number a positive coefficient for a RHS variable
indicates an impact towards the decrease of the spatial dispersion of
population. In other words, a positive coefficient indicates an effect
towards less uniform city sizes that is more dispersion of city sizes.
The latter is indicative of enhancement of agglomeration economies
because of the expansion of digital technologies.

\begin{table}[!htbp] \centering 
  \caption{OLS estimation of equation (3)\label{ols.global}} 
  \label{} 
\small 
\begin{tabular}{@{\extracolsep{1pt}}lcccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{4}{c}{\textit{Dependent variable:}} \\ 
\cline{2-5} 
\\[-1.8ex] & \multicolumn{4}{c}{Pareto exponent 2000-18} \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4)\\ 
\hline \\[-1.8ex] 
 Internet users per 100 hab. (log) & 0.001$^{***}$ &  &  &  \\ 
  & (0.0001) &  &  &  \\ 
  Broadband users per 100 hab. (log) &  & $-$0.001$^{**}$ &  &  \\ 
  &  & (0.0002) &  &  \\ 
  Mobile phone users per 100 hab. (log) &  &  & 0.0003 &  \\ 
  &  &  & (0.0003) &  \\ 
  Fixed phone users per 100 hab. (log) &  &  &  & $-$0.00003 \\ 
  &  &  &  & (0.0001) \\ 
  Population density (log) & $-$1.378 & $-$1.060 & $-$1.505 & $-$1.892 \\ 
  & (1.870) & (1.985) & (1.896) & (2.007) \\ 
  Government expenditure (% GDP) & 0.0002 & 0.003$^{***}$ & 0.001 & 0.001 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  Trade (\% of GDP) & 0.0002$^{***}$ & $-$0.0001 & 0.0001 & 0.0001 \\ 
  & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\ 
  Non agriculture value added (\% GDP) & 0.0001 & 0.001 & $-$0.0001 & $-$0.0002 \\ 
  & (0.001) & (0.001) & (0.001) & (0.001) \\ 
  GDP growth & 0.001$^{***}$ & 0.001$^{***}$ & 0.001$^{**}$ & 0.001$^{**}$ \\ 
  & (0.0003) & (0.0004) & (0.0004) & (0.0004) \\ 
  GDP per capita (log) & $-$0.020$^{***}$ & $-$0.023$^{***}$ & $-$0.019$^{***}$ & $-$0.016$^{**}$ \\ 
  & (0.007) & (0.006) & (0.007) & (0.008) \\ 
  Population (log) & 1.300 & 0.955 & 1.390 & 1.785 \\ 
  & (1.873) & (1.985) & (1.899) & (2.011) \\ 
  Constant & $-$18.795 & $-$13.969 & $-$19.948 & $-$25.516 \\ 
  & (26.314) & (27.886) & (26.670) & (28.261) \\ 
 \hline \\[-1.8ex] 
Country fixed effects & Yes & Yes & Yes & Yes \\ 
Yearly fixed effects & Yes & Yes & Yes & Yes \\ 
Observations & 844 & 757 & 865 & 867 \\ 
Adjusted R$^{2}$ & 0.987 & 0.988 & 0.986 & 0.986 \\ 
Residual Std. Error & 0.516 & 0.497 & 0.540 & 0.540 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{4}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 & \multicolumn{4}{r}{Robust std. errors in parenthesis} \\ 
\end{tabular} 
\end{table}

Table \ref{ols.global} and \ref{2sls.global} report the estimation
results based on the UN urban agglomeration data. Starting with the
former, we note that an agglomerative effect is only detected for the
internet users and while a marginally opposing result can be seen for
broadband users as indicated by the significant coefficients in columns
(1) and (2). For the telephony variables the estimation of \ref{eq3} did
not yield statistically significant coefficients. Before discussing
further these results and the effects of the other control variables, we
need to highlight that the main challenge of estimating equation
\ref{eq3} is the potential endogenous nature of the share of internet
users which might prevent us from being able to infer a truly causal
effect. Endogeneity might be an issue here as spatial structure, which
is proxied by the \color{blue} Pareto exponent, \color{black} might be
affected by another source, which also affects internet penetration. For
instance, economic development might affect the concentration of
population in large cities and at the same time enable more people to go
online. If we do not address this issue, the coefficient for the main
variable of interest will capture potential effects that internet
penetration has on spatial structure, but also potential reverse
causality effects that spatial structure might generate on internet
penetration. To overcome this potential problem, Table \ref{2sls.global}
reports estimates of equation \ref{eq3} using two-stage least squares
(2SLS) with instrumental variables (IVs). The latter are variables which
are correlated with our endogenous variables, but do not influence
current spatial structure. Such an approach will enable us to estimate
the causal effect -- if any -- of the internet and digital
communications on spatial structure. At a first stage, our endogenous
variables are regressed against the IVs. Then, the predicted values of
the endogenous variables based on the IVs and the other control
variables are used instead of the endogenous variable to estimate eq.
\ref{eq3}. A significant effect will verify the causal impact of the
internet and digital communication usage on spatial structure.
